prolfqua&colon; A Comprehensive R-package for Proteomics Differential Expression Analysis

### 2022-10-22 inst/MyArticle/paper.Rmd
Once more review paper and fix bibliography

---
title: prolfqua&colon; A Comprehensive R-package for Proteomics Differential Expression Analysis
author: 
  - Witold E. Wolski $^{\dagger\ddagger}$^[[Functional Genomics Center Zurich (FGCZ)](https://fgcz.ch) - [University of Zurich](https://uzh.ch)/[ETH Zurich](https://ethz.ch), Winterthurerstrasse 190, CH-8057 Zurich, Switzerland. $^\ddagger$[Swiss Institute of Bioinformatics (SIB), Quartier Sorge - Batiment Amphipole, 1015 Lausanne, Switzerland](https://www.sib.swiss/) $^\dagger$Correspondence&colon; \texttt{wew@fgcz.ethz.ch}]
  - Paolo Nanni$^{\ast}$
  - Jonas Grossmann$^{\ast\ddagger}$
  - Maria d'Errico$^{\ast\ddagger}$
  - Ralph Schlapbach$^{\ast}$
  - Christian Panse$^{\ast\ddagger}$
output: 
  bookdown::pdf_document2:
      toc: true
      toc_appendix: false
      toc_bib: false
      keep_tex: true
      citation_package: natbib
date: "`r format(Sys.time())`"
bibliography: prolfqua.bib
vignette: >
  %\VignetteIndexEntry{prolfqua&colon; A Comprehensive R-package for Proteomics Differential Expression Analysis} 
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
abstract: |
    Mass spectrometry is widely used for quantitative proteomics studies, relative protein quantification, and differential expression analysis of proteins. There is a large variety of signal processing software and analysis tools. Nevertheless, there is a need for a modular, easy-to-use application programming interface in R that transparently supports a variety of well principled statistical procedures to make applying them to proteomics data, comparing and understanding their differences easy. The `r BiocStyle::Githubpkg("fgcz/prolfqua")` package integrates essential steps of the mass spectrometry-based differential expression analysis workflow: quality control, data normalization, protein aggregation, statistical modeling, hypothesis testing, and sample size estimation. The package makes integrating new data formats easy. It can be used to model simple experimental designs with a single explanatory variable and complex experiments with multiple factors and hypothesis testing. The implemented methods allow sensitive and specific differential expression analysis. Furthermore, the package implements benchmark functionality that can help to compare data acquisition, data preprocessing, or data modeling methods using a gold standard dataset. The application programmer interface of `r BiocStyle::Githubpkg("fgcz/prolfqua")`  strives to be clear,  predictable, discoverable, and consistent to make proteomics data analysis easy and exciting. Finally, the `r BiocStyle::Githubpkg("fgcz/prolfqua")` R package is available on GitHub https://github.com/fgcz/prolfqua, distributed under the MIT license. It runs on all platforms supported by the [R free software environment for statistical computing and graphics](https://www.r-project.org/).

@@

By plotting the $TPR$ versus the $FPR$ we obtain the receiver operator characteristic curve (ROC curve) [@pROC_2011]. The area under the curve ($AUC$) or partial areas under the curve ($pAUC$), at various values of the $FPR$, are measures of performance derived from the ROC curve. Using these measures, we can compare the performances of the statistics or the various models examined and test if these differences are statistically significant.

@@

colnames(xM) <- c("prolfqua functions", "model")
knitr::kable(tibble::as_tibble(xM), caption = "Prolfqua functions that can be used to fit various models.",
             format = "latex", booktabs = TRUE) |>
    kableExtra::kable_styling(latex_options = "scale_down") |> 
    kableExtra::add_footnote(label = "in development")

@@

Since only technical replicates are available for each dilution, essential sources of variation present in every experiment, such as the biochemical and biological, are not measured. Therefore, the IonStar dataset (see Methods) captures only the variance from the chromatography, electro-spray, and mass spectrometric measurements. Furthermore, this benchmark with a parallel group design does not allow for benchmarking models with interactions. Unfortunately,  other benchmark datasets have similar constraints [@edwards2015cptac, @o2018proteome]. At the same time, we could not analyze the benchmark dataset [@wessels2012comprehensive] with a full factorial design with an up-to-date version of MaxQuant.
Thus, while we can extrapolate some of the results obtained to more realistic datasets, we need to be careful not to over-interpret. Specifically, the observed variances will be higher in more realistic experiments, and the power will be lower for the same sample sizes. Furthermore, the proportion of missing observations in real-life datasets might also be higher.

@@

The R-package `r BiocStyle::Biocpkg("proDA")`, `r BiocStyle::Biocpkg("msqrob2")` and `r BiocStyle::Biocpkg("prolfqua")` (see Methods) model the missing data directly, while MSstats imputes the data using an accelerated failure model. Despite imputation, `r BiocStyle::Biocpkg("MSstats")` estimates group differences for less proteins and does not achieve a higher $pAUC_{10}$. Furthermore, Figure \@ref(fig:figure3) Panel C shows that when using `r BiocStyle::Biocpkg("MSstats")` the proportion of false discoveries might be very high even when filtering for a low $FDR$.

We obtained difference estimates for all proteins, as shown in Figure \@ref(fig:figure3) Panel A when using the probabilistic dropout model (`r BiocStyle::Biocpkg("proDA")`), the hurdle model (`r BiocStyle::Biocpkg("msqrob2")`), and when combining the results of the linear model with contrast estimates in the presence of missing data (prolfqua_merged). Furthermore, the performance of the scaled $p$-values or the $t$-statistics is comparable with that of the other methods (Figure \@ref(fig:figure3) Panel B). All three models perform similarly well when benchmarked against the IonStar/MaxQuant, IonStar/FragPipe, or CPTAC dataset (see Supplement, Benchmark CPTAC). We tested if the $pAUC_{10}$ differences for the three methods were statistically significant, and none of them were at a significance level of $0.4$ (see Supplement, Testing ROC curves). Furthermore, for the IonStar/MaxQuant dataset, msqrob2 performed best, proDA performed best for the CPTAC/MaxQuant data, and prolfqua_merged for the IonStar/FragPipe dataset. Also, the FDR estimates (Figure \@ref(fig:figure3) Panel C) are comparable for all three methods.

Since, the API of `r BiocStyle::Biocpkg("proDA")` and the produced outputs are similar to those of prolfqua, we are planning to integrate the `r BiocStyle::Biocpkg("proDA")` package as an additional modeling option into `r BiocStyle::Githubpkg("fgcz/prolfqua")` (see Figure \@ref(fig:ContrastUML)), to make it easy for application developers to use both.

We focused our benchmark on comparing the statistical modeling methods using three different scores while we fixed the pre-processing steps. However, there are other equally or even more important parameters of a protein quantification pipeline [@frohlich2022benchmarking]. One of them is the normalization of the intensities within the samples to remove systematic differences [@pursiheimo2015optimization]. The method used to infer protein intensities from peptide intensities is an additional important factor [@Grossmann2010]. For instance, the original `r BiocStyle::Biocpkg("proDA")` publication uses MaxLFQ [@cox2014accurate] protein estimates. However, when using MaxLFQ intensities reported by MaxQuant, the $pAUC_{10}$ is significantly lower ($pAUC_{10}(\textrm{t-statistics}) = 66%$) compared with results obtained when protein abundances are estimated from peptide abundances using Tukey's median polish ($pAUC_{10}(\textrm{t-statistics}) = 75%$). Last but not least, the software [@Cox2008MaxQuant; @yu2020fast], identifying proteins and generating the quantification values, significantly contributes to the entire pipeline's performance, altering the number of identified proteins and the sensitivity and specificity of the differential expression analysis. In the supplement we show the benchmarking results for `r BiocStyle::Biocpkg("proDA")` and `r BiocStyle::Githubpkg("fgcz/prolfqua")` when the IonStar dataset is prepossessed with the MSFragger/IonQuant [@yu2020fast] quantification pipeline (v15). Although the number of proteins identified with two peptides is slightly lower than with MaxQuant (3929 vs. 4046), the $pAUC_{10}$ is significantly higher ($81%$ vs. $72%$).

@@

`r BiocStyle::Githubpkg("fgcz/prolfqua")` is a feature-rich, object-oriented R package to analyze quantitative mass spectrometric data with simple or complex experimental designs. While other R packages for differential expression analysis of proteins typically only implement one modeling approach, `r BiocStyle::Githubpkg("fgcz/prolfqua")` supports various types of models (see Figure 2 and Table 3). Furthermore, the contrast specification is explicit and consistent for all models and allows for testing interactions. Furthermore, the modular design of `r BiocStyle::Githubpkg("fgcz/prolfqua")` enables adding new features, e.g., generalized linear models to model the presence or absence information of a protein, or robust linear models, in the future. Furthermore, the developed framework is flexible enough to integrate other modeling methods, e.g., the probabilistic dropout model [@bioxrvproDA2020] or accurate variance estimation [@zhu2020deqms]. Hence, `r BiocStyle::Githubpkg("fgcz/prolfqua")` enables the implementation of applications where the user can select an alternative normalization method, protein abundance estimation method, or differential expression analysis algorithm. Furthermore, with minimal adaptations, this R package can analyze different quantitative proteomics data (e.g., labeling-based TMT-, PRM- DIA-data).

>>>>

---
title: prolfqua&colon; A Comprehensive R-package for Proteomics Differential Expression Analysis
author: 
  - Witold E. Wolski $^{\dagger\ddagger}$^[[Functional Genomics Center Zurich (FGCZ)](https://fgcz.ch) - [University of Zurich](https://uzh.ch)/[ETH Zurich](https://ethz.ch), Winterthurerstrasse 190, CH-8057 Zurich, Switzerland. $^\ddagger$[Swiss Institute of Bioinformatics (SIB), Quartier Sorge - Batiment Amphipole, 1015 Lausanne, Switzerland](https://www.sib.swiss/) $^\dagger$Correspondence&colon; \texttt{wew@fgcz.ethz.ch}]
  - Paolo Nanni$^{\ast}$
  - Jonas Grossmann$^{\ast\ddagger}$
  - Maria d'Errico$^{\ast\ddagger}$
  - Ralph Schlapbach$^{\ast}$
  - Christian Panse$^{\ast\ddagger}$
output: 
  bookdown::pdf_document2:
      toc: true
      toc_appendix: false
      toc_bib: false
      keep_tex: true
      citation_package: natbib
date: "`r format(Sys.time())`"
bibliography: prolfqua.bib
vignette: >
  %\VignetteIndexEntry{prolfqua&colon; A Comprehensive R-package for Proteomics Differential Expression Analysis} 
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
abstract: |
    Mass spectrometry is widely used for quantitative proteomics studies, relative protein quantification, and differential expression analysis of proteins. There is a large variety of signal processing software and analysis tools. Nevertheless, there is a need for a modular, easy-to-use application programming interface in R that transparently supports a variety of well principled statistical procedures to make applying them to proteomics data, comparing and understanding their differences easy. The `r BiocStyle::Githubpkg("fgcz/prolfqua")` package integrates essential steps of the mass spectrometry-based differential expression analysis workflow: quality control, data normalization, protein aggregation, statistical modeling, hypothesis testing, and sample size estimation. The package makes integrating new data formats easy. It can be used to model simple experimental designs with a single explanatory variable and complex experiments with multiple factors and hypothesis testing. The implemented methods allow sensitive and specific differential expression analysis. Furthermore, the package implements benchmark functionality that can help to compare data acquisition, data preprocessing, or data modeling methods using a gold standard dataset. The application programmer interface of `r BiocStyle::Githubpkg("fgcz/prolfqua")`  strives to be clear,  predictable, discoverable, and consistent to make proteomics data analysis application development easy and exciting. Finally, the `r BiocStyle::Githubpkg("fgcz/prolfqua")` R package is available on GitHub https://github.com/fgcz/prolfqua, distributed under the MIT license. It runs on all platforms supported by the [R free software environment for statistical computing and graphics](https://www.r-project.org/).

@@

By plotting the $TPR$ versus the $FPR$ we obtain the receiver operator characteristic curve (ROC curve) [@pROC2011]. The area under the curve ($AUC$) or partial areas under the curve ($pAUC$), at various values of the $FPR$, are measures of performance derived from the ROC curve. Using these measures, we can compare the performances of the statistics or the various models examined and test if these differences are statistically significant.

@@

colnames(xM) <- c("prolfqua functions", "model")
knitr::kable(tibble::as_tibble(xM), caption = "prolfqua functions that can be used to fit various models.",
             format = "latex", booktabs = TRUE) |>
    kableExtra::kable_styling(latex_options = "scale_down") |> 
    kableExtra::add_footnote(label = "in development")

@@

The IonStar dataset (see Methods) captures only the variance from the chromatography, electro-spray, and mass spectrometric measurements since only technical replicates are available for each dilution. Therefore essential sources of variation typically present in other experiments, such as the biochemical and biological, are not measured. Furthermore, this dataset with a parallel group design does not allow for benchmarking models with interactions. However, other benchmark datasets have similar constraints [@edwards2015cptac, @o2018proteome]. Furthermore, we could not analyze the benchmark dataset [@wessels2012comprehensive] with a complete factorial design with an up-to-date version of MaxQuant. Thus, while we can extrapolate some of the results obtained to more realistic datasets, we need to be careful not to over-interpret our findings. Specifically, the observed variances will be higher in datasets with biological replicates, and the power will be lower for the same number of samples. Furthermore, the proportion of missing observations in real-life datasets might be higher or distributed differently in groups.

@@

The R-package `r BiocStyle::Biocpkg("proDA")`, `r BiocStyle::Biocpkg("msqrob2")` and `r BiocStyle::Biocpkg("prolfqua")` do not impute missing data but integrate them into the statistical model, while MSstats filters and imputes the data using an accelerated failure model. Despite imputation, `r BiocStyle::Biocpkg("MSstats")` estimates group differences for less proteins and does not achieve a higher $pAUC_{10}$. Furthermore, Figure \@ref(fig:figure3) Panel C shows that when using `r BiocStyle::Biocpkg("MSstats")` the proportion of false discoveries might be very high even when filtering for a low $FDR$.

We obtained difference estimates for all proteins, as shown in Figure \@ref(fig:figure3) Panel A when using: a) the probabilistic dropout model (`r BiocStyle::Biocpkg("proDA")`), b) the hurdle model (`r BiocStyle::Biocpkg("msqrob2")`), and c) when combining the results of the linear model with contrast estimates in the presence of missing data (prolfqua_merged). Furthermore, the performance of the scaled $p$-values or the $t$-statistics is comparable among these three methods (Figure \@ref(fig:figure3) Panel B). We tested if there is no difference between the $pAUC_{10}$ for all three methods and did not reject this hypothesis at a significance level of $0.4$ (see Supplement, Testing ROC curves). Furthermore, all three models perform similarly well when examined using the IonStar/MaxQuant, IonStar/FragPipe, or CPTAC dataset (see Supplement, Benchmark CPTAC, and Benchmark FragPipe). While, for the IonStar/MaxQuant dataset, msqrob2 performed best, proDA performed best for the CPTAC/MaxQuant data, and prolfqua_merged for the IonStar/FragPipe dataset. Also, the FDR estimates (Figure \@ref(fig:figure3) Panel C) are comparable for all three methods.

Since, the API of `r BiocStyle::Biocpkg("proDA")` and the produced outputs are similar to those of prolfqua, we are integrating the `r BiocStyle::Biocpkg("proDA")` package as an additional modeling option into `r BiocStyle::Githubpkg("fgcz/prolfqua")` (see Figure \@ref(fig:ContrastUML)).

We focused our benchmark on comparing the statistical modeling methods using three different scores while we fixed the pre-processing steps. However, some of these steps are of utmost significance when performing a differential expression analysis [@frohlich2022benchmarking]. One of them is the normalization of the intensities within the samples to remove systematic differences [@pursiheimo2015optimization]. The method used to infer proteins from peptide identifications [@nesvizhskii2005interpretation] and protein abundances from peptide abundances is an additional important factor [@Grossmann2010]. For instance, the original `r BiocStyle::Biocpkg("proDA")` publication uses MaxLFQ [@cox2014accurate] protein estimates. However, when using MaxLFQ intensities reported by MaxQuant, the $pAUC_{10}$ is lower ($pAUC_{10}(\textrm{t-statistics}) = 66%$) compared with results obtained when protein abundances are estimated from peptides abundances using Tukey's median polish ($pAUC_{10}(\textrm{t-statistics}) = 72%$). Last but not least, the software [@Cox2008MaxQuant; @yu2020fast], used to identify and quantify proteins, significantly contributes to the entire pipeline's performance, altering the number of identified proteins and the sensitivity and specificity of the differential expression analysis. In the supplement we show the benchmarking results for `r BiocStyle::Biocpkg("proDA")` and `r BiocStyle::Githubpkg("fgcz/prolfqua")` when the IonStar dataset is prepossessed with the FragPipe v15 [@yu2020fast] quantification pipeline. Although the number of proteins identified with two peptides is slightly lower than with MaxQuant (3929 vs. 4046), the pAUC_{10} is significantly higher ($81%$ vs. $72%$).

@@

`r BiocStyle::Githubpkg("fgcz/prolfqua")` is a feature-rich, object-oriented R package to analyze quantitative mass spectrometric data with simple or complex experimental designs. While other R packages for differential expression analysis of proteins typically only implement one modeling approach, `r BiocStyle::Githubpkg("fgcz/prolfqua")` supports various types of models (see Figure \@ref(fig:ContrastUML) and Table \@ref(tab:prolfquaModels)). Furthermore, the contrast specification is explicit and consistent for all models and allows for testing interactions. The modular design of `r BiocStyle::Githubpkg("fgcz/prolfqua")` enables adding new features, e.g., generalized linear models to model the presence or absence information of a protein, or robust linear models, in the future. Furthermore, the developed framework can integrate other modeling methods, e.g., the probabilistic dropout model [@bioxrvproDA2020] or accurate variance estimation [@zhu2020deqms]. Hence, `r BiocStyle::Githubpkg("fgcz/prolfqua")` enables the implementation of applications where the user can select an alternative normalization method, protein abundance estimation method, or differential expression analysis algorithm. Furthermore, this R package can analyze other types of quantitative proteomics data,e.g., labeling-based TMT-, PRM- DIA-data.
